---
title: "Introduction to RFmstate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RFmstate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**RFmstate** fits clock-reset cause-specific random survival forests for
acyclic, non-recurrent multistate processes. For each transient state,
competing exits are modeled by separate forests. Patient/profile
entry-conditioned state probabilities are assembled from predicted cumulative
hazards by semi-Markov entry-mass and sojourn convolution. The package also
provides calendar-time Aalen-Johansen point estimates as a covariate-free
descriptive baseline. The supported one-row-per-subject contract uses a common
initial state, one recorded entry per state, baseline covariates, right
censoring, and competing exits; it does not support left truncation, recurrent
visits, directed cycles, or time-dependent covariates.

The package provides:

- State space and transition structure definition
- Wide-to-long data conversion for counting-process format
- Cause-specific random forest fitting per origin state
- Entry-conditioned state probabilities via clock-reset semi-Markov convolution
- Aalen-Johansen calendar-time point estimation (covariate-free baseline)
- Per-transition feature importance
- Genuine edge OOB concordance and patient-level cross-validated IPCW scoring
- Comprehensive visualizations

## Quick Start

### 1. Define the Multistate Structure

```{r define-states}
library(RFmstate)

# Use the built-in clinical trial structure
ms <- clinical_states()
print(ms)
```

Or define a custom structure with any number of states (3 or more):

```{r custom-states, eval=FALSE}
# A simple 3-state illness-death model
ms_simple <- define_multistate(
  state_names = c("Healthy", "Sick", "Dead"),
  absorbing = "Dead",
  transitions = list(
    Healthy = c("Sick", "Dead"),
    Sick = c("Dead")
  )
)

# A 4-state model with recovery
ms_recovery <- define_multistate(
  state_names = c("Healthy", "Sick", "Recovered", "Dead"),
  absorbing = "Dead",
  transitions = list(
    Healthy = c("Sick", "Dead"),
    Sick = c("Recovered", "Dead"),
    Recovered = c("Dead")
  )
)
```

This pipeline works with any defined multistate structure.

### 2. Simulate Data

```{r simulate}
set.seed(42)
dat <- sim_clinical_data(n = 300, structure = ms)
head(dat)
```

### 3. Prepare Multistate Data

Convert wide-format data to long format:

```{r prepare}
msdata <- prepare_data(
  data = dat, id = "ID", structure = ms,
  time_map = list(
    Responded = "time_Responded",
    Unresponded = "time_Unresponded",
    Stabilized = "time_Stabilized",
    Progressed = "time_Progressed",
    Death = "time_Death"
  ),
  censor_col = "time_censored",
  covariates = c("age", "sex", "BMI", "treatment")
)
print(msdata)
```

### 4. Aalen-Johansen Nonparametric Baseline

Compute nonparametric (covariate-free) transition probability estimates:

```{r aj}
aj <- aalen_johansen(msdata)
print(aj)
```

```{r aj-plot, fig.cap="State occupation probabilities from Aalen-Johansen estimator"}
plot(aj, type = "state_occupation")
```

```{r aj-hazard, fig.cap="Nelson-Aalen cumulative hazards"}
plot(aj, type = "cumulative_hazard")
```

### 5. Fit Random Forest Model

```{r fit}
fit <- rfmstate(
  msdata,
  covariates = c("age", "sex", "BMI", "treatment"),
  num.trees = 200,
  seed = 42
)
print(fit)
```

### 6. Model Summary

```{r summary}
s <- summary(fit)
```

### 7. Feature Importance

```{r importance, fig.cap="Feature importance per transition"}
imp <- importance(fit)
print(imp)
plot(imp, type = "barplot")
```

```{r importance-heat, fig.cap="Feature importance heatmap"}
plot(imp, type = "heatmap")
```

### 8. Predict for New Patients

```{r predict, fig.cap="Predicted state occupation for two patient profiles"}
newdata <- data.frame(
  age = c(50, 70),
  sex = c(0, 1),
  BMI = c(24, 32),
  treatment = c(1, 0)
)

prediction_horizon <- min(fit$max_duration_by_origin)
pred <- predict(fit, newdata = newdata,
                times = seq(0, prediction_horizon, length.out = 37))

# Plot for patient 1 (young, treated)
plot(pred, type = "state_occupation", subject = 1)

# Plot for patient 2 (older, untreated)
plot(pred, type = "state_occupation", subject = 2)
```

### 9. Diagnostics

```{r diagnostics}
diag <- diagnose(fit)
print(diag)
```

```{r diag-concordance, fig.cap="Concordance index per transition"}
plot(diag, type = "concordance")
```

Full-state Brier scores require patient-level cross-validation and refitting;
they are never assembled from incompatible edge-level OOB predictions:

```{r diag-cv, eval=FALSE}
cv_diag <- diagnose(fit, method = "cv", folds = 5,
                    eval_times = seq(0, prediction_horizon * 0.8,
                                     length.out = 9))
plot(cv_diag, type = "brier")
```

### 10. Transition Diagram

```{r diagram, fig.cap="Transition diagram with event counts"}
plot_transition_diagram(ms, msdata)
```

## Methodology

### Separate probability constructions

RFmstate forests use duration since fresh entry into the current state. Their
predicted cause-specific cumulative hazards are combined by semi-Markov
entry-mass and sojourn convolution on a validated regular duration grid. The
output is an entry-conditioned state-occupation array, not a general Markov
$P(s,t)$ matrix.

The Aalen-Johansen baseline is separate: it uses calendar-time risk sets and a
product integral from the recorded common study origin.

### Aalen-Johansen Estimator (Nonparametric Baseline)

The Aalen-Johansen (AJ) estimator is the nonparametric generalization of the
Kaplan-Meier estimator to multistate models under the Markov assumption. It
estimates hazard increments from the data directly via the Nelson-Aalen
formula:

$$d\hat{A}_{hj}(u) = \frac{dN_{hj}(u)}{Y_h(u)}$$

where $dN_{hj}(u)$ counts the observed $h \to j$ transitions at time $u$ and
$Y_h(u)$ is the number at risk in state $h$ just before time $u$. This
provides population-level transition probabilities without covariate
adjustment and serves as a covariate-free baseline in the package.

### Random Forest Multistate Approach

For covariate-adjusted predictions, we decompose the multistate model into
per-origin-state competing risks problems:

1. **For each transient state** $h$, identify all outgoing transitions
2. **Fit a cause-specific RSF**: For each destination state $j$, fit a
   random survival forest treating transition $h \to j$ as the event of
   interest and all other transitions as censored
3. **Extract cumulative hazards**: Use ranger's predicted cumulative hazard
   for each declared edge on its observed duration support
4. **Assemble by convolution**: combine state-entry masses, origin-state
   sojourn survival, and cause-specific next-exit distributions in topological
   order

This approach leverages the flexibility of random forests to capture
nonlinear covariate effects and interactions while maintaining the
interpretability of the approved acyclic, non-recurrent multistate scope.
Analytic, probability-invariant, and grid-refinement checks validate the
numerical approximation without clipping or row normalization.

### Diagnostics

- **OOB Error**: Out-of-bag prediction error from the random forest ensemble
- **Edge OOB C-index**: genuine ranger OOB concordance for each binary
  cause-specific endpoint; it is not a full-model validation score
- **Full-state Brier/IBS**: IPCW scores from patient-level held-out complete
  pipeline predictions

## References

- Aalen, O.O. & Johansen, S. (1978). An empirical transition matrix for
  non-homogeneous Markov chains based on censored observations.
  *Scandinavian Journal of Statistics*, 5(3), 141-150.
- Ishwaran, H. et al. (2008). Random survival forests.
  *Annals of Applied Statistics*, 2(3), 841-860.
- Putter, H., Fiocco, M. & Geskus, R.B. (2007). Tutorial in biostatistics:
  Competing risks and multi-state models. *Statistics in Medicine*, 26,
  2389-2430.
